### 2020-02-08 doc/icfp20/counterpoint.tex
integrate unisonCheck into intervalCheck

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Here is an example of first species counterpoint, where the lower line
stands for the cantus firmus and the upper line represents the
counterpoint\footnote{The cantus firmus is part of a German song
  called ``Froschgesang'' (Frog's song).}:

[Frog score]

We represent such music in the following way: we keep the first and
last bars as separate elements, and encode the middle bars as a vector
of pitch-interval pairs \texttt{(p , i)}, where \texttt{p} is a note in the
cantus firmus, and \texttt{i} is the distance between \texttt{p} and the
corresponding counterpoint note.
For instance, the main body of the above example is represented as
follows:

@@

\paragraph{Beginning}
The beginning of the music must be a perfect consonance,
that is, it must be a unison, a 5th, or an 8th.
In our implementation, this is checked by the predicate
\texttt{checkBeginning}, which is defined as follows:

@@

\noindent The datatype \texttt{BeginningError} is a kind of
``type error'' designed specifically for counterpoint.
The first constructor \texttt{not158} tells us that the music
begins with a wrong interval, whereas the second constructor
\texttt{tooShort} says that the whole music is empty.
With this datatype, we can obtain an informative error message
when we accidentally break the rule at the beginning of the music.
Thus, what we are doing here can be seen as writing a
customized typechecker.

\paragraph{Intervals}
In the main body of the music, we must ensure that vertical intervals
are harmonically pleasing.
all intervals must be consonant.
That is, we should only use 3rds, 5ths, 6ths, and 8ths.
This rule can be encoded as the \texttt{checkIntervals} predicate,
which again uses a new datatype \texttt{IntervalError} representing
the presence of a dissonant interval:

\begin{alltt}
data IntervalError : Set where
  dissonant : Interval \(\rightarrow\) IntervalError

intervalCheck : Interval \(\rightarrow\) Maybe IntervalError
intervalCheck i = if isConsonant i then nothing else just (dissonant i)

checkIntervals : List PitchInterval \(\rightarrow\) List IntervalError
checkIntervals = mapMaybe (intervalCheck \(\circ\) proj\(\sb{\mathtt{2}}\))
\end{alltt}

\paragraph{Motion}

\begin{alltt}
data Motion : Set where
contrary : Motion
parallel : Motion
similar  : Motion
oblique  : Motion
\end{alltt}

For every pair of two adjacent intervals, the cantus firmus and
counterpoint lines must not move in the same direction if the second
interval is a perfect one (i.e., unison, 5th, or 8th).
This is because such motion would draw too much attention to the
resulting interval.

@@

checkMotion : List PitchInterval → List MotionError
checkMotion = mapMaybe (uncurry motionCheck) ∘ pairs  
\end{alltt}

\paragraph{Unisons}
Unisons should not occur in the main body of the music.
The reason is that they make it harder to distinguish between
the two melodic lines.

\begin{alltt}
data UnisonError : Set where
  unison : Pitch \(\rightarrow\) UnisonError

unisonCheck : PitchInterval \(\rightarrow\) Maybe UnisonError
unisonCheck (p , i) = if (i == per1) then just (unison p) else nothing

checkUnison : List PitchInterval \(\rightarrow\) List UnisonError
checkUnison = mapMaybe unisonCheck
\end{alltt}

\paragraph{Ending}
The ending of the music must be a \emph{cadence}, a progression
of two intervals that gives rise to a sense of resolution.

>>>>

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Figure~\ref{fig:frog} is an example of first species counterpoint,
where the lower line stands for the cantus firmus and the upper line
represents the counterpoint\footnote{The cantus firmus is part of a
  German song called ``Froschgesang'' (Frog's song).}:

Among different options of how to encode such music, we adopt
the following strategy.
First, we represent each bar as a pitch-interval pair \texttt{(p ,  i)},
where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
distance between \texttt{p} and the corresponding counterpoint note.
Second, we keep the first and last bars as separate elements, and
represent the middle bars (often referred to as the \emph{main body})
as a list of pitch-interval pairs.
For instance, the main body of the above example is represented as
follows:

@@

\paragraph{Beginning}
The beginning of the music should express perfection.
Therefore, the first interval must be a unison, a 5th, or an 8th.
In our formalization, we implement this rule as a function
\texttt{checkBeginning}, which reports an error \texttt{not158 i}
when the first interval \texttt{i} is an invalid one.

@@

\paragraph{Intervals}
The main body of the music should maintain harmonical consonance
and independence of melodic lines.
This means, all constituent intervals must be chosen from the 3rd,
5th, 6th, and 8th.
We encode this rule as the \texttt{checkIntervals} function,
which returns a list of errors correponding to the occurrences of
dissonant intervals and unisons.

\begin{alltt}
data IntervalError : Set where
  dissonant : Interval \(\rightarrow\) IntervalError
  unison    : Pitch \(\rightarrow\) IntervalError

intervalCheck : PitchInterval \(\rightarrow\) Maybe IntervalError
intervalCheck (p , i) with isConsonant i | isUnison i
... | false | _    = just (dissonant i)
... | _     | true = just (unison p)
... | _     | _    = nothing

checkIntervals : List PitchInterval \(\rightarrow\) List IntervalError
checkIntervals = mapMaybe intervalCheck
\end{alltt}

\paragraph{Motion}
The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As shown in Figure~\ref{fig:motion}, there are four kinds of motion,
and among them, parallel and similar motion makes it harder to
distinguish between two voices, especially when the second interval
is a perfect one.
To avoid this problematic motion, we define the \texttt{checkMotion}
function, which lists all invalid motion approaching a perfect interval.

@@

checkMotion : List PitchInterval \(\rightarrow\) List MotionError
checkMotion = mapMaybe (uncurry motionCheck) \(\circ\) pairs  
\end{alltt}

\paragraph{Ending}
The ending of the music should express relaxation.
This means, the last interval must be chosen from the unison and
the 8th, which are both considered very stable.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
We encode these rules as the \texttt{checkEnding} function;
note that it returns the \texttt{tooShort} error when the music does
not have enough number of bars to form a valid ending.

\begin{alltt}
data EndingError : Set where
  not18    : PitchInterval \(\rightarrow\) EndingError
  not27    : PitchInterval \(\rightarrow\) EndingError
  tooShort : List PitchInterval \(\rightarrow\) EndingError

endingCheck : PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe EndingError
endingCheck pi1@(pitch p , i) (pitch q , interval 0)  = 
  if ((p + 1 \(\equiv\sp{\mathtt{b}}\) q) \(\wedge\) (i == min3)) then nothing else just (not27 pi1)
endingCheck pi1@(pitch p , i) (pitch q , interval 12) =
  if ((q + 2 \(\equiv\sp{\mathtt{b}}\) p) \(\wedge\) (i == maj6) \(\vee\) (p + 1 \(\equiv\sp{\mathtt{b}}\) q) \(\wedge\) (i == min10))
  then nothing
  else just (not27 pi1)
endingCheck pi1               pi2                     =
  just (not18 pi2)

checkEnding : List PitchInterval \(\rightarrow\) PitchInterval \(\rightarrow\) Maybe EndingError
checkEnding []        \_ = just (tooShort [])
checkEnding (p :: []) q = endingCheck p q
checkEnding (p :: ps) q = checkEnding ps q
\end{alltt}
